{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-11 paper/si.tex
wip NG working on journal revision

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfold on two time scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ R &\ S\  \cr
		D &\ T &\ P\ \cr
	} . \label{eq:prisoner}
\end{align}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play many repeated games with randomly changing interaction partners.

@@

The long time scale describes the evolutionary dynamics. 
Here, players are allowed to update their strategies based on the payoffs they yield. 
We model these strategy updates with a pairwise comparison process~\citep{traulsen2007pairwise}.
This process assumes that at regular time intervals, one player is randomly selected from the population.
We refer to this player as the `learner' (\learner). 
The learner is then given an opportunity to update its strategy. 
There are two possibilities for how this update may occur. 
With probability $\mu$, the player's strategy mutates randomly. 
In that case, the player's new strategy is drawn uniformly from the space of all reactive strategies~$[0,1]^3$.  
With probability $1\!-\!\mu$, the player compares itself with a neighbor. 
In that case, the player randomly picks another individual from the population (referred to as the `role model', \rolemodel). 
The learner adopts the role model's strategy with a probability \(\varphi\), given by
\begin{equation} \label{Eq:rho}
    \varphi(\pi_{\learner}, \pi_{\rolemodel}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_\rolemodel - \pi_\learner)}}.
\end{equation}
The parameter \(\beta\) is the selection strength.
It determines how important payoff differences are for the learner's decision to imitate the role model. 
The variables $\pi_\rolemodel$ and $\pi_\learner$ refer to the relevant payoffs of the role model and the learner, respectively. 
The exact value of these payoffs depend on the players' memory. 
We say players have {\it perfect memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' expected payoffs (across all rounds and across all possible co-players). 
We say players have {\it limited memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' realized payoff in the very last round of the game with their very last interaction partner. 
In addition, we consider several model extensions in which individuals have memory capacities in between these two extremes. 
We provide a detailed description of these different settings and the resulting payoffs in Section~\ref{section:analyticalresults}.\\

@@

In the following, we discuss our different memory settings in more detail. 
We discuss six cases explicitly. In these cases, updating occurs
({\it i})~based on average payoffs based on all interactions (perfect memory), 
({\it ii})~based on the last round of one interaction (limited memory), 
({\it iii})~based on the last round of two interactions,
({\it iv})~based on the last two rounds of one interaction,
({\it v})~based on the last two rounds of two interactions, and
({\it v})~based on the average payoff of one interaction.
In each case, we consider the case that there are only two strategies present in the population (a resident and a mutant strategy). 
We first derive how likely it is that a learner (of any type) assigns a given payoff $\pi_\learner$ to itself, and a payoff of $\pi_\rolemodel$ to the role model. 
This allows us to derive explicit expressions for $\lambda^-_k/\lambda^+_k$, and hence for the mutant's fixation probability according to Eq.~\eqref{eq:fixation_probability}.
Based on these expressions we can characterize under which conditions cooperation is stochastically stable.

@@

{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}

@@

Here, individuals update their strategies based on the expected payoffs, based on all rounds and all possible interaction partners. 

@@

\right]

@@

@@

~\\
\noindent
After these preparations, consider now a population with \(k\) mutants and \(N - k\) residents, whose strategies we denote by \(\strategy_\mutant=(y_\mutant,p_\mutant, q_\mutant)\) and \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\), respectively.
Assuming that population members are matched randomly (or equivalently, that they interact with all other population members), the resulting expected payoffs of residents and mutants are
\begin{equation} \label{Eq:ExpPay}
  \begin{array}{lcrcr}
  \displaystyle \pi_\resident(k)& = &\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \pi(\strategy_\resident,\strategy_\resident)	&+	&\displaystyle\frac{k}{N-1}\cdot \pi(\strategy_\resident,\strategy_\mutant),\\[0.5cm]
  \displaystyle \pi_\mutant(k)& = &\displaystyle\frac{N-k}{N-1}\cdot \pi(\strategy_\mutant,\strategy_\resident) &+	&\displaystyle\frac{k-1}{N-1}\cdot \pi(\strategy_\mutant,\strategy_\mutant).\\
  \end{array}
\end{equation}
The number of mutants in the population decreases in a single time step if a mutant is chosen to be the learner and adopts the strategy
of a resident. 
Similarly, it increases if a resident is the learner and adopts the strategy of a mutant. 
The respective transition probabilities are 
\begin{align*}
  \lambda^-_k \!=\! \frac{N-k}{N}\frac{k}{N-1}\varphi\big(\pi_\mutant(k),\,\pi_\resident(k)\big) \quad \text{ and } \quad \lambda^+_k \!=\!\frac{k}{N}\frac{N-k}{N-1}\varphi\big(\pi_\resident(k),\,\pi_\mutant(k)\big).
\end{align*}
For $\varphi$ as defined by Eq.~\eqref{Eq:rho}, the ratio of these two transition probabilities simplifies to 
\begin{equation} \label{eq:LambdaRatio}
 \frac{\lambda^-_k}{\lambda^+_k} \!=\!  \frac{\varphi\big(\,\pi_\mutant(k),~\pi_\resident(k)\,\big)}{\varphi\big(\,\pi_\resident(k),~\pi_\mutant(k)\,\big)} 
 =e^{-\beta\big(\pi_\mutant(k)-\pi_\resident(k)\big)}.
\end{equation}
Based on these ratios for each $k$, we can compute the mutant's fixation probability by Eq.~\eqref{eq:fixation_probability}.

@@

~\\
{\bf Stochastic stability of cooperation.}
As an application of this formalism, we can compute when cooperation is stochastically stable in the perfect-information setting. 
To this end, suppose there is only a single mutant, $k\!=\!1$. The residents adopt Generous Tit-for-Tat, $\gtft = (1, 1, q)$ and the mutant adopts $\alld\! = \! (0,0,0)$. When two GTFT players interact, the resulting average distribution according to Eq.~\eqref{Eq:vAverage} simplifies to
\begin{align*}
\mathbf{v}(\gtft,\gtft) = (1, 0, 0, 0).
\end{align*}
On the other hand, if \alld{} interacts with \gtft, the respective probabilities become,
\begin{align*}
  \mathbf{v}(\alld,\gtft) = \big(0,~0,~1\!-\!\delta\!+\!\delta q,~\delta(1\! -\! q)\big).
\end{align*}
Based on Eq.~\eqref{Eq:ExpPay}, we can compute the strategies' expected payoffs as
\begin{align*}
\pi_\gtft =  \frac{N\!-\!2}{N-1} (b - c)  \!-\! \frac{1}{N-1}\big(1\!-\!\delta\!+\!\delta q\big)c
\quad \text{and} \quad
\pi_\alld  = \big(1\!-\!\delta\!+\!\delta q\big)b.
\end{align*}
As a consequence, we can calculate the corresponding ratio of transition probabilities according to Eq.~\eqref{eq:LambdaRatio},
\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b+\frac{c}{N-1})- \frac{N-2}{N-1}(b-c) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation}\label{Eq:Condition_PerfectMemory}
q<1-\frac{1}{\delta}\cdot \frac{b+(N\!-\!1)c}{(N\!-\!1)b+c}.
\end{equation}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta \!>\! \big(b+(N\!-\!1)c\big)/\big((N\!-\!1)b\!+\!c\big)$.
In particular, in the limit of large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is stochastically stable if  \(q \!<\! 1\! -\! c/(\delta b)\). 
The minimum continuation probability for such a strategy to exist is $\delta \!>\! c/b$. 
In this way, we recover the classical conditions for cooperation to be feasible under direct reciprocity~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}.

@@

{\bf Computing the distribution of last-round payoffs.}
The case of perfect payoff memory is straightforward to handle; here, every player gets the expected payoff with certainty. 
In comparison, computing transition probabilities for the case of limited payoff memory is more elaborate. 
Here, we need to consider the different possible outcomes that both the learner and role model may have experienced in their very last interaction. 
A further complication arises when the learner's last interaction partner happens to be the role model. 
In that case, the learner's and the role model's last payoff will be correlated (e.g., if the learner got the sucker's payoff of $S$, the role model's payoff is $T$ with certainty). 
To treat the case of limited memory analytically, let $\mathcal{U} \!=\! \{R,S,T,P\}$ be the set of possible one-shot payoffs. 
Since the game ends after round $T$ with probability $\delta^{T}\left(1-\delta\right)$, if two players use reactive strategies $\strategy_1\!=\!(y_1, p_1, q_1$)  and $\strategy_2\!=\!(y_2,p_2,q_2)$, then, by definition, the probability that the first player receives a payoff of $u\!\in\!\mathcal{U}$ in the final round of the game is given by Eq.~\eqref{Eq:vAverage}. To make the strategies explicit, we denote this probability by $v_{u}\!\left(\strategy_{1},\strategy_{2}\right)$.

@@

~\\
\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
After these preparations, let us again consider the corresponding population setup, with $N\!-\!k$ residents with strategy  \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\) and $k$ mutants with strategy  \(\strategy_\mutant = (y_\mutant, p_\resident, q_\mutant)\).
At each step of the evolutionary process we choose a learner and a role model. 
The learner compares the performance of its strategy by comparing its own last one-shot payoff with the last one-shot payoff of the role model. 
In the following, we assume either the learner or the role model is a resident and that the other player is a mutant (otherwise it is certain that the number of mutants does not change). 
There are two major cases to consider.

@@

\begin{enumerate}
\item The learner and the role model have their last respective interaction with each other. This happens with probability $1/(N\!-\!1)$. In that case, there are four possible cases for their joint final payoffs, $(u_\resident,u_\mutant)\!\in\!\mathcal{U}_F\!:=\!\big\{(R,R),\,(S,T),\,(T,S),\,(P,P)\big\}$. These outcomes follow the distribution $\mathbf{v}(\strategy_\resident,\strategy_\mutant)$ (Eq.~\eqref{Eq:vAverage}).  
\item The learner's last interaction was not with the role model, with probability $(N\!-\!2)/(N\!-\!1)$. In this case, there are four different subcases, depending on whether the resident's last interaction partner was a mutant or a resident, and depending on whether the mutant's last interaction partner was a mutant or a resident. As a result, the resident's last one-shot payoff is distributed according to $\mathbf{v}(\strategy_\resident,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\resident,\strategy_\resident)$; the mutant's last payoff is distributed according to $\mathbf{v}(\strategy_\mutant,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\mutant,\strategy_\resident)$, respectively. 
\end{enumerate}

@@

\noindent
The first term on the right side corresponds to the case that the learner and
the role model happened to be matched directly for their last interaction. 
In that case, only those payoff pairs can occur that are feasible in a direct interaction. 
That is, it needs to be the case that $(u_\resident,u_\mutant)\in \mathcal{U}_F$, as represented by the respective indicator function. 
The probability that the number of mutants increase or decreases by one is now given by
\begin{equation}
\begin{array}{l}
\displaystyle \lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_\resident,u_\mutant\,\in\,\mathcal{U}} x(u_\resident,u_\mutant)\cdot \varphi(u_\resident,u_\mutant), \\[0.5cm]
\displaystyle \lambda^-_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_\resident,u_\mutant \,\in\, \mathcal{U}} x(u_\resident,u_\mutant)\cdot \varphi(u_\mutant,u_\resident).
\end{array}
\end{equation}

@@

{\bf Motivation.} 
In the limited-memory setting, individuals take into account the outcome of one round, and of one interaction. 
This setting can be generalized such that an individual
considers \(m\) rounds and of \(n\) interactions. Here, we discuss the case that
the update depends on the last round of \(n\) interactions.

At each step of the evolutionary process, we consider the role model's and the learner's last \(n\) matches. 
We need to define the probability that for each of
the matches they are paired with a mutant, with a resident or with each other.
We assume that each pair is unique, such that the learner and the role model
can be matched at most once, which is a reasonable assumption in large populations.
The case of \(n\!=\!1\) corresponds to the previous setting of limited memory. 
There, we have seen that there are five possible combinations to consider. 
As we increase $n$, the number of possible combinations increases non-linearly; 
for a graphical illustration, see Fig.~\ref{fig:matching_tree}. 
In the following, we study the case of \(n=2\), such that the learner takes into account the players' final payoffs of two interactions.\\

@@

We assume that player the resident receives the payoff \(u_{\resident1}\) with their first interaction and \(u_{\resident 2}\) with their second.
Similarly, the mutant receives payoffs \(u_{\mutant 1}\) and \(u_{\mutant 2}\). 
Let \(x(u_{\resident1} u_{\resident1}, u_{\mutant 1} u_{\mutant 2})\) denote the probability that the resident and the mutant received payoffs  \(u_{\resident1}\), \(u_{\resident2}\) and \(u_{\mutant 1}\), \(u_{\mutant 2}\) in the last round of their respective last two interactions.
This probability can be computed as follows, 
\begin{equation} \label{Eq:xTwoInteractions} 
  \setlength{\arraycolsep}{0.5cm}
  \footnotesize
  \hspace*{-0.8cm} 
  \begin{array}{lll}
  \multicolumn{3}{l}{x(u_{\resident1}u_{\resident 2}, u_{\mutant 1}u_{\mutant 2})= \displaystyle \frac{1}{N\!-\!1}\cdot v_{u_{\resident1}}(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_{\resident1},u_{\mutant 1})\in \mathcal{U}_F} \cdot A }\\[0.4cm]

@@

As before, we can use this formalism to compute when cooperation is stochastically stable. For a single \alld{} mutant in a \gtft{} resident population, we obtain the following non-zero probabilities \(x(u_{\resident1} u_{\resident 2}, u_{\mutant 1},u_{\mutant 2})\)

@@

We note that unlike the case of limited payoff memory, here the exact magnitude of the benefit and the cost of cooperation do have some effect on the transition probabilities. 

@@

{\bf General setup.} As our next setting, we consider a setup in which players update their strategies based on the last two rounds of their last game. 
Here, additional complications may arise because some games may only last for one round if $\delta\!<\!1$.
In the following, we abstract from these complications.  
We shall assume that learners only update their strategies when both the learner's and the role model's last game had at least two rounds. 
If either the learner or the role model only engaged in a one-round game in their previous interaction, we assume the learner decides to keep its strategy. 
As a result, in the following, we accordingly re-interpret some of our quantities. 
For example, $\lambda^+_k$ is no longer the (unconditional) probability that the number of mutants increases by one after a learner and a role model have been chosen.
Instead, we interpret it as the (conditional) probability that the number of mutants increases, given that the learner and the role model happened to interact at least for two rounds during their respective last interaction. 
This new interpretation of transition probabilities affects the time scale of fixation. 
After all, we ignore all time steps in which no update occurred because a player's last game only lasted for a single round.
However, our re-interpretation does not affect the ratio $\lambda^-_k/\lambda^+_k$, which is the key quantity we consider.\\

@@

Based on these payoff probabilities, we compute the ratio

@@

\subsection{Updates based on average payoff of one repeated game}\label{section:one_interaction_full_memory}

@@

{\bf Motivation.}

\nikoleta{CH to write a few senteces}

\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
We revisit the population setup, consisting of \(k\) mutants with the strategy
\(\strategy_\mutant = (y_\mutant, p_\mutant, q_\mutant)\) and \(N-k\) residents
with strategies \(\strategy_\resident = (y_\resident, p_\resident,
q_\resident)\). In each step of the evolutionary process, we select a learner and
a role model. The learner or the role model can be either a resident, and the
other player is a mutant. We consider that the learner and the
model update based on a single repeated game with a member of the population.
As mentioned previously, two major cases must be considered:

>>>>

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfold on two time scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ R &\ S\  \cr
		D &\ T &\ P\ \cr
	} . \label{eq:prisoner}
\end{align}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\strategy\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play repeated games with all other population members.

@@

The long time scale describes the evolutionary dynamics. 
Here, players are allowed to update their strategies based on the payoffs they yield. 
We model these strategy updates with a pairwise comparison process~\citep{traulsen2007pairwise}.
This process assumes that at regular time intervals, one player is randomly selected from the population.
We refer to this player as the `learner' (\learner). 
The learner is given an opportunity to update its strategy. 
There are two possibilities for how this update may occur. 
With probability $\mu$, the player's strategy mutates randomly. 
In that case, the player's new strategy is drawn uniformly from the space of all reactive strategies~$[0,1]^3$.  
With probability $1\!-\!\mu$, the player compares itself to another population member. 
To this end, the player randomly picks another individual from the population (referred to as the `role model', \rolemodel). 
The learner adopts the role model's strategy with a probability \(\varphi\), given by
\begin{equation} \label{Eq:rho}
    \varphi(\pi_{\learner}, \pi_{\rolemodel}) = \frac{1}{1\!+\! e^{-\!\beta (\pi_\rolemodel - \pi_\learner)}}.
\end{equation}
The parameter \(\beta\!\ge\!0\) is the selection strength.
It determines how important payoff differences are for the learner's decision to imitate the role model. 
The variables $\pi_\rolemodel$ and $\pi_\learner$ refer to the payoffs of the role model and the learner, respectively. 
The exact value of these payoffs depend on the players' memory. 
We say players have {\it perfect memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' expected payoffs (across all rounds and across all possible co-players). 
We say players have {\it limited memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' realized payoff in the very last round of the game with their very last interaction partner. 
In addition, we consider several model extensions in which individuals have memory capacities in between these two extremes. 
We provide a detailed description of these different settings and the resulting payoffs in Section~\ref{section:analyticalresults}.\\

@@

In the following, we discuss our different memory settings in more detail. 
We discuss six cases explicitly. In these cases, updating occurs
({\it i})~based on average payoffs based on all interactions (perfect memory), 
({\it ii})~based on the last round of one interaction (limited memory), 
({\it iii})~based on the last round of two interactions,
({\it iv})~based on the last two rounds of one interaction,
({\it v})~based on the last two rounds of two interactions, and
({\it v})~based on the average payoff of one interaction.
In each case, we assumet there are only two strategies present in the population (a resident and a mutant strategy). 
We first derive how likely it is that a learner (of any type) assigns a given payoff $\pi_\learner$ to itself, and a payoff of $\pi_\rolemodel$ to the role model. 
This allows us to derive explicit expressions for $\lambda^-_k/\lambda^+_k$, and hence for the mutant's fixation probability according to Eq.~\eqref{eq:fixation_probability}.
Based on these expressions we can characterize under which conditions cooperation is stochastically stable.

@@

{\bf Computing the players' expected payoffs.}

@@

Here, individuals update their strategies based on their expected payoffs, taking into account all rounds and all possible interaction partners. 

@@

\right].

@@

The second player's payoff can be computed analogously.

@@

~\\
\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
After these preparations, consider now a population with \(k\) mutants and \(N - k\) residents, whose strategies we denote by \(\strategy_\mutant=(y_\mutant,p_\mutant, q_\mutant)\) and \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\), respectively.
Assuming that population members are matched randomly (or equivalently, that they interact with all other population members), the resulting expected payoffs of residents and mutants are
\begin{equation} \label{Eq:ExpPay}
  \begin{array}{lcrcr}
  \displaystyle \pi_\resident(k)& = &\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \pi(\strategy_\resident,\strategy_\resident)	&+	&\displaystyle\frac{k}{N-1}\cdot \pi(\strategy_\resident,\strategy_\mutant),\\[0.5cm]
  \displaystyle \pi_\mutant(k)& = &\displaystyle\frac{N-k}{N-1}\cdot \pi(\strategy_\mutant,\strategy_\resident) &+	&\displaystyle\frac{k-1}{N-1}\cdot \pi(\strategy_\mutant,\strategy_\mutant).\\
  \end{array}
\end{equation}
The number of mutants in the population decreases in a single time step if a mutant is chosen to be the learner and adopts the strategy
of a resident. 
Similarly, it increases if a resident is the learner and adopts the strategy of a mutant. 
The respective transition probabilities are 
\begin{align*}
  \lambda^-_k \!=\! \frac{N-k}{N}\frac{k}{N-1}\varphi\big(\pi_\mutant(k),\,\pi_\resident(k)\big) \quad \text{ and } \quad \lambda^+_k \!=\!\frac{k}{N}\frac{N-k}{N-1}\varphi\big(\pi_\resident(k),\,\pi_\mutant(k)\big).
\end{align*}
For $\varphi$ as defined by Eq.~\eqref{Eq:rho}, the ratio of these two transition probabilities simplifies to 
\begin{equation} \label{eq:LambdaRatio}
 \frac{\lambda^-_k}{\lambda^+_k} \!=\!  \frac{\varphi\big(\,\pi_\mutant(k),~\pi_\resident(k)\,\big)}{\varphi\big(\,\pi_\resident(k),~\pi_\mutant(k)\,\big)} 
 =e^{-\beta\big(\pi_\mutant(k)-\pi_\resident(k)\big)}.
\end{equation}
Based on these ratios for each $k$, we can compute the mutant's fixation probability by Eq.~\eqref{eq:fixation_probability}.

@@

~\\
{\bf Stochastic stability of cooperation.}
As an application of this formalism, we can compute when cooperation is stochastically stable in the perfect-information setting. 
To this end, suppose there is a single mutant, $k\!=\!1$. The residents adopt Generous Tit-for-Tat, $\gtft = (1, 1, q)$ and the mutant adopts $\alld\! = \! (0,0,0)$. When two GTFT players interact, the resulting average distribution according to Eq.~\eqref{Eq:vAverage} simplifies to
\begin{align*}
\mathbf{v}(\gtft,\gtft) = (1, 0, 0, 0).
\end{align*}
On the other hand, if \alld{} interacts with \gtft, the respective probabilities become,
\begin{align*}
  \mathbf{v}(\alld,\gtft) = \big(0,~0,~1\!-\!\delta\!+\!\delta q,~\delta(1\! -\! q)\big).
\end{align*}
Based on Eq.~\eqref{Eq:ExpPay}, we can compute the strategies' expected payoffs as
\begin{align*}
\pi_\gtft =  \frac{N\!-\!2}{N-1} (b - c)  \!-\! \frac{1}{N-1}\big(1\!-\!\delta\!+\!\delta q\big)c
\quad \text{and} \quad
\pi_\alld  = \big(1\!-\!\delta\!+\!\delta q\big)b.
\end{align*}
As a consequence, we can calculate the corresponding ratio of transition probabilities according to Eq.~\eqref{eq:LambdaRatio},
\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b+\frac{c}{N-1})- \frac{N-2}{N-1}(b-c) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation}\label{Eq:Condition_PerfectMemory}
q<1-\frac{1}{\delta}\cdot \frac{b+(N\!-\!1)c}{(N\!-\!1)b+c}.
\end{equation}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta \!>\! \big(b+(N\!-\!1)c\big)/\big((N\!-\!1)b\!+\!c\big)$.
In particular, in the limit of large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is stochastically stable if  \(q \!<\! 1\! -\! c/(\delta b)\). 
The minimum continuation probability for such a strategy to exist is $\delta \!>\! c/b$. 
In this way, we recover the classical conditions for cooperation to be feasible under direct reciprocity~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}.

@@

{\bf Computing the distribution of last-round payoffs.}
The case of perfect payoff memory is straightforward to handle; here, every player gets the expected payoff with certainty. 
In comparison, computing transition probabilities for the case of limited payoff memory is more elaborate. 
Here, we need to consider the different possible outcomes that both the learner and role model may have experienced in their very last interaction. 
A further complication arises when the learner's last interaction partner happens to be the role model. 
In that case, the learner's and the role model's last payoff will be correlated (e.g., if the learner got the sucker's payoff of $S$, the role model's payoff is $T$ with certainty). 
To treat the case of limited memory analytically, let $\mathcal{U} \!=\! \{R,S,T,P\}$ be the set of possible one-shot payoffs. 
Since the game ends after round $\tau$ with probability $\delta^{\tau}\left(1-\delta\right)$, if two players use reactive strategies $\strategy_1\!=\!(y_1, p_1, q_1$)  and $\strategy_2\!=\!(y_2,p_2,q_2)$, then, by definition, the probability that the first player receives a payoff of $u\!\in\!\mathcal{U}$ in the final round of the game is given by Eq.~\eqref{Eq:vAverage}. To make the strategies explicit, we denote this probability by $v_{u}\!\left(\strategy_{1},\strategy_{2}\right)$.

@@

~\\
\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
After these preparations, let us again consider the corresponding population setup, with $N\!-\!k$ residents with strategy  \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\) and $k$ mutants with strategy  \(\strategy_\mutant = (y_\mutant, p_\mutant, q_\mutant)\).
At each step of the evolutionary process we choose a learner and a role model. 
The learner compares the performance of its strategy by comparing its own last one-shot payoff with the last one-shot payoff of the role model. 
In the following, we assume either the learner or the role model is a resident, and that the other player is a mutant (otherwise the number of mutants does not change). 
There are two major cases to consider.

@@

\begin{enumerate}
\item The learner and the role model have their last respective interaction with each other. This happens with probability $1/(N\!-\!1)$. In that case, there are four possible cases for their joint final payoffs, $(u_\resident,u_\mutant)\in\mathcal{U}_F\!:=\!\big\{(R,R),\,(S,T),\,(T,S),\,(P,P)\big\}$. By Eq.~\eqref{Eq:vAverage}, these four outcomes follow the distribution~$\mathbf{v}(\strategy_\resident,\strategy_\mutant)$.
\item The learner's last interaction was not with the role model, with probability $(N\!-\!2)/(N\!-\!1)$. In this case, there are four different subcases, depending on whether the resident's last interaction partner was a mutant or a resident, and depending on whether the mutant's last interaction partner was a mutant or a resident. As a result, the resident's last one-shot payoff is distributed according to $\mathbf{v}(\strategy_\resident,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\resident,\strategy_\resident)$; the mutant's last payoff is distributed according to $\mathbf{v}(\strategy_\mutant,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\mutant,\strategy_\resident)$, respectively. 
\end{enumerate}

@@

\noindent
The first line on the right side corresponds to the case that the learner and
the role model happened to be matched directly for their last interaction. 
In that case, only those payoff pairs can occur that are feasible in a direct interaction. 
That is, it needs to be the case that $(u_\resident,u_\mutant)\in \mathcal{U}_F$, as represented by the respective indicator function. 
The second and third line of \eqref{eq:Chi} summarize the four possible subcases that can occur when the learner and the role model were not directly matched for their last interaction.

Overall, we now obtain the following expressions for the probability that the number of mutants increases or decreases by one, \begin{equation}
\begin{array}{l}
\displaystyle \lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_\resident,u_\mutant\,\in\,\mathcal{U}} x(u_\resident,u_\mutant)\cdot \varphi(u_\resident,u_\mutant), \\[0.5cm]
\displaystyle \lambda^-_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_\resident,u_\mutant \,\in\, \mathcal{U}} x(u_\resident,u_\mutant)\cdot \varphi(u_\mutant,u_\resident).
\end{array}
\end{equation}

@@

{\bf Motivation.} 
In the limited-memory setting, individuals take into account the outcome of one round, and of one interaction. 
This setting can be generalized such that an individual
considers \(m\) rounds and \(n\) interactions. Here, we discuss the case that
the update depends on the last round of \(n\) interactions.

At each step of the evolutionary process, we consider the role model's and the learner's last \(n\) matches. 
We need to define the probability that for each of
the matches they are paired with a mutant, with a resident or with each other.
We assume that each pair is unique, such that the learner and the role model
can be matched at most once, which is a reasonable assumption in large populations.
The case of \(n\!=\!1\) corresponds to the previous setting of limited memory. 
There, we have seen that there are five possible combinations to consider. 
As we increase $n$, the number of possible combinations increases non-linearly; 
for a graphical illustration, see Supplementary~Fig.~\ref{fig:matching_tree}. 
In the following, we study the case of \(n=2\), such that the learner takes into account the players' final payoffs of two interactions.\\

@@

We assume the resident receives the payoff \(u_{\resident1}\) with their first interaction and \(u_{\resident 2}\) with their second.
Similarly, the mutant receives payoffs \(u_{\mutant 1}\) and \(u_{\mutant 2}\). 
Let \(x(u_{\resident1} u_{\resident1}, u_{\mutant 1} u_{\mutant 2})\) denote the probability that the resident and the mutant received payoffs  \(u_{\resident1}\), \(u_{\resident2}\) and \(u_{\mutant 1}\), \(u_{\mutant 2}\) in the last round of their respective last two interactions.
This probability can be computed as follows, 
\begin{equation} \label{Eq:xTwoInteractions} 
  \setlength{\arraycolsep}{0.5cm}
  \footnotesize
  \hspace*{-0.8cm} 
  \begin{array}{lll}
  \multicolumn{3}{l}{x(u_{\resident1}u_{\resident 2}, u_{\mutant 1}u_{\mutant 2})= \displaystyle \frac{1}{N\!-\!1}\cdot v_{u_{\resident1}}(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_{\resident1},u_{\mutant 1})\in \mathcal{U}_F} \cdot A }\\[0.4cm]

@@

As before, we can use this formalism to compute when cooperation is stochastically stable. For a single \alld{} mutant in a \gtft{} resident population, we obtain the following non-zero probabilities \(x(u_{\resident1} u_{\resident 2}, u_{\mutant 1},u_{\mutant 2})\),

@@

Unlike in the case of limited payoff memory, here the exact magnitude of the benefit and the cost of cooperation do have some effect on the transition probabilities. 

@@

{\bf General setup.} As our next setting, we assume players update their strategies based on the last two rounds of their last game. 
Here, additional complications may arise because some games may only last for one round if $\delta\!<\!1$.
In the following, we abstract from these complications.  
We shall assume that learners only update their strategies when both the learner's and the role model's last game had at least two rounds. 
If either the learner or the role model only engaged in a one-round game in their previous interaction, we assume the learner decides to keep its strategy. 
As a result, in the following, we accordingly re-interpret some of our quantities. 
For example, $\lambda^+_k$ is no longer the (unconditional) probability that the number of mutants increases by one after a learner and a role model have been chosen.
Instead, we interpret it as the (conditional) probability that the number of mutants increases, given that the learner and the role model happened to interact at least for two rounds during their respective last interaction. 
This new interpretation of transition probabilities affects the time scale of fixation. 
After all, we ignore all time steps in which no update occurred because a player's last game only lasted for a single round.
However, our re-interpretation does not affect the ratio $\lambda^-_k/\lambda^+_k$, which is the key quantity we consider.\\

@@

Based on these probabilities, we compute the ratio

@@

\subsection{Updates based on the average payoff of one repeated game}\label{section:one_interaction_full_memory}

@@

{\bf Motivation.} As our last setting, we consider the case that players update their strategies based on an (entire) repeated game against a single co-player. In terms of cognitive complexity, this setting is in between the case of perfect memory and the case considered in Section~\ref{section:m_two_n_one}, where players only take into account two rounds of one repeated game.\\
